PODD: An Ontology-driven Data Repository for Collaborative Phenomics Research

### 2012-03-02 icadl2010/slides/slides.tex
initial files from http://projects.arcs.org.au/svn/podd/trunk/doc/publications/

\title[PODD -- Ontology-driven Repository]{PODD: An Ontology-driven Data Repository for Collaborative Phenomics Research}

@@

\institute[UQ]{
    The eResearch Lab, School of ITEE\\The University of Queensland\\
    }
\vspace{+80pt}
\date[IBS Workshop 2010]{\footnotesize IBS Phenomics Data and Informatics Workshop - Canberra 2010}

@@

\section{Introduction}
\begin{frame}{Motivation}
\begin{block}{Challenges in phenomics research data management}
    \begin{itemize}
        \item Data is huge
        \begin{itemize}
            \item APN: estimated 1.8TB + 30\% growth/year
            \item TPA: 0.5TB/week $\sim 25$ TB/year
        \end{itemize}
        \pause
        \item New platforms/processes/technologies emerge
        \pause
        \item Data needs \textbf{\emph{context}}
        \begin{itemize}
            \item Scientific, administrative \& other metadata
        \end{itemize}
    \end{itemize}
\end{block}
\pause
\begin{block}{Repositories for the management of data}
    \begin{itemize}
      \item Not all questions answered
    \end{itemize}
\end{block}
\end{frame}

@@

1. The volume of data. Phenomics research generates large amounts of
data. For example, APN, the Australian Phenomics Network, is
estimated to generate 1.8TB of data in the first year with 30\%
growth. That is more than 7TB over 3 years. And APN is only one
research center. We still have APPF, which operates 2 centers. The
large volume makes the storage and distribution much more difficult.

2. Phenomics is a new area. As a result, new imaging platforms,
technologies and processes emerge all the time. A data repository
needs to be constructed in such a way that it can cater for future
platforms and technologies.

3. Data needs context. 7TB of data isn't very useful if it just sits
there and collects dust. To make use of the data, we need to
surround them with context, the metadata, that describe the raw data
in a meaningful way. Metadata can have dimensions, such as
scientific and administrative. They can help researchers to discover
data, integrate data and analyze data is a vital part of a data
repository.

There are quite a few databases for genomics and protein research,
such as Emsembl and UniProt. Unfortunately, not all the requirements
for phenomics data management are met. For example, genomic data is
relatively well defined, which isn't the case for phenomics.
Phenomics data may come off from a variety of imaging and measuring
platforms and the appropriate metadata needs to be captured.

@@

\begin{frame}{Clients \& Collaborators}
\begin{block}{Australian Integrated Biological Science Facilities}
    \begin{itemize}
        \item<2-> Australian Plant Phenomics Facility (APPF)
            \begin{itemize}
                \item High-throughput (TPA) \& high-resolution (HRPPC) centers
            \end{itemize}
        \item<3-> Australian Phenomics Network (APN)
            \begin{itemize}
                \item Mouse models, \emph{deep} imaging and measuring platforms
            \end{itemize}
        \item<4-> Atlas of Living Australia (ALA)
            \begin{itemize}
                \item Biodiversity data collection \& management
            \end{itemize}
    \end{itemize}
\end{block}
~\\
\begin{figure}[ht]
\centering
\onslide<2->\subfigure[]{
\includegraphics[width=20mm]{conveyor.png}
}
\onslide<2->\subfigure[]{
\includegraphics[width=20mm]{sample.png}
}
\onslide<3->\subfigure[]{
\includegraphics[width=20mm]{mouse.png}
}
\onslide<4->\subfigure[]{
\includegraphics[width=20mm]{organism.png}
}
\end{figure}
\end{frame}

@@

Australian Plant Phenomics Facility has two research centers.

@@

The High Resolution Plant Phenomics Center, the HRPPC, is located
here in Canberra. The center focuses on deep phenotyping of plants
using a number of equipments. Some of them operate on single plants
in controlled environments, others operate on fields.

The Australian Phenomics Network (APN) is a collaborative research
facility established to provides mouse models for human and animal
diseases. For example, APN uses digital slide scanner to perform
histopathology and organ pathology imaging and analysis.

The Atlas of Living Australia is developing a biodiversity data
management system for Australia. Collaboration with ALA is also one
of PODD's goals.

\begin{frame}{Data Management Requirements}
\begin{overprint}
\onslide<1>
\begin{block}{Data capturing}
\begin{tabularx}{\linewidth}{p{35mm}X}
Flow Cytometry & FACS data \\
Histopathology & Zeiss slide images \\
Plant imaging & Lemnatec images, Flourogroscan images, 3D imaging \\
Infrared imaging & FLIR images \\
Chemical measurements & Chlorophyll content, Stomatal conductance \\
Visual observation & Manual reports\\
$\cdots$ & $\cdots$
\end{tabularx}
\end{block}

\onslide<2>
\begin{block}{Data generation}
\begin{tabularx}{\linewidth}{p{30mm}X}
Project & Project proposal, project plan \\
Investigation & Objectives, design \\
Materials & Lines/genotypes, samples, growth conditions \\
Devices & Specs, settings, versions \\
Processes & Workflows, protocols, variations \\
Measurements & Data, images \\
Analysis & Observations, results \\
$\cdots$ & $\cdots$
\end{tabularx}
\end{block}

\onslide<3>
\begin{block}{Data management}
\begin{tabularx}{\linewidth}{p{50mm}l}
Data distribution & ? \\
Data sharing & ? \\
Data publishing & ? \\
Access control & ? \\
Archival \& versioning & ? \\
Structure \& metadata & ? \\
Data discovery \& analysis & ?
\end{tabularx}
\end{block}
\end{overprint}
\end{frame}

One of the requirements of PODD is to support data capturing from
various imaging platforms.

As you can see on this slide. there are quite a number of different
ways of imaging and observing the subject. The challenge here is,
besides capturing the raw image files and observations, capture
sufficient metadata.

Another requirement is data generation. Data and metadata can be
generated on different levels and at different times. All this
information needs to be recorded in the repository.

Last but not least, there is a need for the effective management of
data. Tasks such as distribution, publishing, access control and
versioning belong to this category. As far as we know, there is no
data repository that handles these tasks for phenomics research.

\begin{frame}{PODD: an ontology-driven repository}
\begin{block}{Goals}
    \begin{itemize}
        \item Supports data captured by different platforms
        \item Supports different data formats
        \item Supports effective data management
        \begin{itemize}
            \item Metadata, distribution, access control, discovery, etc.
        \end{itemize}
    \end{itemize}
\end{block}
\pause
\begin{block}{Approach}
    \begin{itemize}
        \item Define domain ontologies -- focus of this talk
        \item Design an appropriate architecture
        \item Develop an ontology-aware repository
    \end{itemize}
\end{block}
\end{frame}

@@

Ontologies are a formal representation of a particular application
domain. Besides being a controlled vocabulary that define important
concepts and relations, an ontology can define complex constraints
on these concepts and relations, making it a very expressive
conceptualization of the domain.

@@

\begin{frame}{PODD: The High-level Architecture}
 \vspace{-6mm}
 \begin{figure}[t]
  \begin{center}
  \includegraphics[trim=48mm 20mm 10mm 35mm, clip, width=130mm]{architecture.pdf}
   \end{center}
  \label{fig}
 \end{figure}
\end{frame}

Here you see the high-level architecture of the PODD repository. On
the basis are a few components. Fedora Commons is the underlying
repository that we use to store all the metadata information. iRODS
is a federated data storage software that can be used by Fedora
Commons. ANDS provides an iRODS-based service called Data Fabric,
which we will use in PODD to host the data captured.

Ontologies will be stored in Fedora Commons as well as in a RDF
triple store to enable data discovery using SPARQL queries and full
text search on metadata. We also use databases to store information
about users, access rights and some other things.

The data access layer works with ontologies and objects and is
responsible for serializing and deserializing objects from the
underlying storage.

The various services and the web interface is guarded by a security
layer that controls access based on fine-grained user access rights.

@@

\section{Related Work}
\begin{frame}{Related Work}
\begin{block}{FuGe -- Functional Genomics Experiment}
    \begin{itemize}
        \item \emph{Material}, \emph{Protocol}, \emph{Data}, etc.
        \item Can be extended to support phenomics
        \item[$\cross$] May be difficult to extend for new concepts
    \end{itemize}
\end{block}
\pause
\begin{block}{OBI -- Ontology for Biomedical Investigations}
    \begin{itemize}
        \item \textit{``An integrated ontology for the description of life-science and clinical
        investigations.''}
        \item Comprehensive: 2,600+ classes, 10,000+ axioms
        \item[$\cross$] Complex, computationally ($\mathcal{SHOIN}(D)$)
    \end{itemize}
\end{block}
\end{frame}

FuGe is a UML and database based conceptual model for functional
genomics experiments. It models important domain concepts such as
material, protocol and data. The model is stored as database
schemas, as a result, FuGe may be difficult to extend.

The OBI ontology - ontology for biomedical investigation is a
comprehensive but complex ontology. It defines domain concepts 
starting from a very high abstract level. As a result, it is 
quite big, containing about 26 hundred classes and 10 thousand
axioms. It is also computationally expensive. For this reason and
that we need to performing reasoning repeatedly in PODD, we chose
not to use OBI as our domain ontology.

\begin{frame}{Related Work}
\begin{block}{Web Ontology Language (OWL)}
\begin{itemize}
    \item Precise, open \& extensible -- exactly what we need!
    \item Provides core vocabularies for expressing complex ontologies -- \emph{data models}
    \item Main language constructs
    \begin{itemize}
        \item Classes, e.g, \texttt{Human}
        \item Predicates , e.g., \texttt{hasParent}
        \item Individuals, e.g., \texttt{Aristotle}
    \end{itemize}
    \item APIs, query engines \& automated reasoners available
\end{itemize}
\end{block}
\end{frame}

@@

\section{The PODD Ontology}
\begin{frame}{The Ontology Approach}
\begin{block}{Benefits}
    \begin{itemize}
    \setlength{\itemsep}{6pt}
    \item<1-> Greater extensibility
        \begin{itemize}
        \item New concepts/relations can be easily added/modified
        \end{itemize}
    \item<2-> Better reuse \& integration
        \begin{itemize}
        \item Ontologies are \emph{open}
        \item Other ontologies can be integrated on multiple levels
        \end{itemize}
    \item<3-> Balance between expressivity \& reasoning complexity
        \begin{itemize}
        \item Formal semantics enables automated query \& analysis
        \item Off-the-shelf tools available
        \end{itemize}
    \end{itemize}
\end{block}
\end{frame}

@@

Firstly, the repository can be more extensible. In PODD we will store all
object information in ontologies. In such a way new concepts and relations
can be added more easily than if the information is stored in a database.

@@

The OWL language is defined with a formal semantics. This enables automated
query answering and analysis. OWL is designed with expressivity and reasoning
complexity in mind so that the core reasoning tasks can be made fully automated. 
As shown in a previous slide, although OWL is a relatively young language,
there are off-the-shelf tools available to use.

\begin{frame}{The PODD Ontology}
\begin{block}{Modeling essentials}
    \begin{itemize}
    \item Domain entities
        \begin{itemize}
        \item Abstract concepts
        \item Concrete objects
        \end{itemize}
    \item Domain entities defined using OWL \emph{ontologies}
    \end{itemize}
\end{block}
\pause
\begin{block}{Modeling in OWL}
    \begin{itemize}
    \item Domain concepts -- \emph{OWL classes}
    \item Inter-concept relations -- \emph{OWL predicates} \& \emph{OWL restrictions}
    \item Concrete domain objects -- \emph{OWL Individuals}
    \item Comments, descriptions -- \emph{OWL annotations}
    \end{itemize}
\end{block}
\end{frame}

\begin{frame}{The PODD Ontology -- An Example}
\begin{example}{The \textbf{\emph{Project}} concept}
    \begin{itemize}
    \item The top-level concept
    \item Constraints on inter-object relations \& attributes -- like everything else
    \end{itemize}
    \pause
    {\small
    \begin{align*}
    Project &\sqsubseteq~ =~ 1~ hasProjectPlan \sqcap \forall hasProjectPlan.ProjectPlan\\
            &\sqsubseteq~ \geq~ 1~ hasInvestigation \sqcap \forall hasInvestigation.Investigation\\
            &\sqsubseteq~ =~ 1~ hasStartDate \sqcap \forall hasStartDate.date\\
            &\sqsubseteq~ \leq~ 1~ hasPublicationDate \sqcap \forall hasPublicationDate.date
    \end{align*}}
\end{example}
\begin{block}{}
    \begin{itemize}
    \item \textcolor{blue}{\textbf{Extensibility}} from inheritance of OWL classes \& predicates
    \end{itemize}
\end{block}
\end{frame}

In this slide I show a simple, incomplete modeling of the top-level concept
in our ontology, the Project. In our model of phenomics, everything starts 
from a research project and all other objects, such as Investigation and
Material are child or descendant objects.

As you will see, we define constraints on attributes and inter-concept relations
of the Project concept. This is how we constrain all of our concepts.

@@

\begin{frame}{The PODD Ontology}
\onslide<2->
\begin{columns}[t]
\column{.5\textwidth}
\begin{itemize}
    \item Inspired by FuGe \& OBI
    \item Aim: define all \emph{essential} domain concepts, attributes \& relations
    \item $\sim30$ classes, \\$\sim 80$ predicates, \\$\sim 200$ Axioms
    \item Faster reasoning, querying, etc.
\end{itemize}
\vfill
\onslide<1->
\column{.5\textwidth}
\vspace{-10mm}
 \begin{figure}[t]
  \begin{center}
  \includegraphics[trim=3mm 26mm 2mm 8mm, clip, height=72mm]{hier.pdf}
   \end{center}
 \end{figure}
\end{columns}
\end{frame}

@@

\begin{frame}{PODD Ontology -- Roles}
\begin{block}{Ontologies \emph{drives} repository functions}
\begin{list}{}{}
\item[\textcolor{blue}{\textbf{Presentation}}] Drives the rendering of the web pages
    \begin{itemize}
    \item Object creation, editing, display, etc.
    \end{itemize}
\item[\textcolor{blue}{\textbf{Management}}] Object (de)serialization to/from ontologies
\item[\textcolor{blue}{\textbf{Validation}}] Ontology reasoning performed
    \begin{itemize}
    \item Validation on object type, cardinality, etc.
    \end{itemize}
\item[\textcolor{blue}{\textbf{Discovery}}] Multiple ways of finding information
    \begin{itemize}
    \item Queries using SPARQL
    \item Searches using Lucene
    \end{itemize}
\end{list}
\end{block}
\end{frame}

@@

\begin{frame}{The PODD Ontology -- Object creation page rendering}
 \begin{figure}[t]
  \begin{center}
  \includegraphics[width=76mm]{create.png}
   \end{center}
 \end{figure}
\end{frame}

\section{Conclusion}
\subsection*{ }
\begin{frame}{Conclusion}
\begin{block}{To recap}
\begin{itemize}
\item Large amounts of data need to be managed
    \begin{itemize}
    \item There is a need for data archival, storage \& discovery
    \end{itemize}
\item Current approaches lacking/inadequate/inflexible
    \begin{itemize}
    \item Emerging processes, platforms, technologies require extensible infrastructure
    \end{itemize}
\item An ontology-driven approach as the foundation of PODD
    \begin{itemize}
    \item Extensible \& open
    \end{itemize}
\end{itemize}
\end{block}
\end{frame}

This slides shows a screen capture of the project creation page.
Note that most of these fields are generated by reading the domain
ontology and figuring out what fields are needed and what their types,
cardinalites are.

For example, in a previous slide we defined that a project must have 
exactly 1 start date attribute and it must be a date. So here we show 
an input box for this attribute and signal that it is mandatory. Note
that lead institution is also a mandatory attribute while the duration
is not. When this project object is saved, the values are checked against
the domain ontology and any errors or omissions will be reported back
to the user.

\begin{frame}{Conclusion}
\begin{block}{Where we are now}
\begin{itemize}
\item PODD ontology stabilizing
\item Development of basic repository functionality
\item Development of PODD web interface
\end{itemize}
\end{block}
\pause
\begin{block}{What's next}
\begin{itemize}
\item Development of batch data import/export processes
\item Development of object discovery services
\item Integrate with AAF federated authentication services
\item \textcolor{gray}{Exposing data for discovery}
\item \textcolor{gray}{Integrating with other data sources}
\end{itemize}
\end{block}
\end{frame}

\begin{frame}
	\frametitle{}
    \begin{center}
\Huge \textsc{\textcolor{blue}{\textbf{Thank You!}}}
\end{center}
\vspace{+20pt}
\begin{block}{Acknowledgment}
Faith Davies, Gavin Kennedy, Jane Hunter\\
eResearch Lab, Schoo of ITEE, UQ
\end{block}
\end{frame}

>>>>

\title[PODD -- An Ontology-driven Repository]{PODD: An Ontology-driven Data Repository for Collaborative Phenomics Research}

@@

\institute[UQ]{
    The eResearch Lab, School of ITEE\\The University of Queensland\\
    }

\vspace{+380pt}
\date[ICADL 2010]{\footnotesize $12^{th}$ International Conference on Asian Digital Libraries -- Gold Coast, Australia}

@@

\section{Introduction}
\begin{frame}{Motivation}
\begin{block}{Challenges in phenomics research data management}
    \begin{itemize}
        \item Data is huge
        \begin{itemize}
%            \item APN: estimated 1.8TB + 30\% growth/year
            \item TPA: 0.5TB/week $\sim 25$ TB/year
        \end{itemize}
        \pause
        \item New platforms/processes/technologies quickly emerge
        \pause
        \item Data needs \textbf{\emph{context}}
        \begin{itemize}
            \item Scientific, administrative \& other metadata
        \end{itemize}
    \end{itemize}
\end{block}
\pause
\begin{block}{Repositories for the management of data}
    \begin{itemize}
      \item Not all questions answered
    \end{itemize}
\end{block}
\end{frame}

@@

1. The volume of data. Phenomics research generates large amounts of
data. For example, The Plant Accelerator, one of the research centers
in Australian Plant Phenomics Facility, is
estimated to generate 25TB of data per year. And TPA is only one
research center. Under APPF there is still the High-resolution Plant
Phenomics Center. The Australian Phenomics Network is yet another center
that works on mouse phenomics research and generates large amounts of data.
The large volume makes the storage and distribution a challenging task.

2. Phenomics is a new area. As a result, new imaging platforms,
technologies and processes emerge all the time. A data repository
needs to be constructed in such a way that it can capture data and
metadata off new platforms and technologies.

3. Data needs context. 25TB of data isn't very useful if it just 
sits there and collects dust. To make use of the data, we need to
surround them with metadata that describe the raw data in a 
meaningful way. There are different kinds of metadata, such as
scientific and administrative. They can help researchers to 
discover data, integrate data and analyze data is a vital part 
of a data repository.

There are quite a few databases for genomics and proteomics research,
such as Emsembl and UniProt. Unfortunately, not all the requirements
for phenomics data management are met. For example, genomic data is
relatively well understood and structured, which isn't the case for 
phenomics. Phenomics data may come off from a variety of imaging and 
measurement platforms and the appropriate metadata needs to be captured.

@@

\begin{frame}{Clients \& Collaborators}
\begin{block}{Australian Integrated Biological Science Facilities}
    \begin{itemize}
        \item<2-> Australian Plant Phenomics Facility (APPF)
            \begin{itemize}
                \item High-throughput (TPA) \& high-resolution (HRPPC) centers
            \end{itemize}
        \item<3-> Australian Phenomics Network (APN)
            \begin{itemize}
                \item Mouse models, \emph{deep} imaging and measuring platforms
            \end{itemize}
        \item<4-> Atlas of Living Australia (ALA)
            \begin{itemize}
                \item Biodiversity information portal
            \end{itemize}
    \end{itemize}
\end{block}
~\\
\begin{figure}[ht]
\centering
\onslide<2->\subfigure[]{
\includegraphics[width=20mm]{conveyor.png}
}
\onslide<2->\subfigure[]{
\includegraphics[width=20mm]{sample.png}
}
\onslide<3->\subfigure[]{
\includegraphics[width=20mm]{mouse.png}
}
\onslide<4->\subfigure[]{
\includegraphics[width=20mm]{organism.png}
}
\end{figure}
\end{frame}

@@

Australian Plant Phenomics Facility operates two research centers.

@@

The High Resolution Plant Phenomics Center, the HRPPC, is located
in Canberra. The center focuses on deep phenotyping of plants
using a variety of equipments. Some of them operate on single 
plants in controlled environments, others operate in fields.

The Australian Phenomics Network (APN) is a collaborative research
facility established to provides mouse models for human and animal
diseases. For example, APN uses digital slide scanners to perform
histopathology and organ pathology imaging and analysis.

The Atlas of Living Australia is developing a biodiversity 
information portal for Australia. Collaboration with ALA is 
also one of PODD's goals.

\begin{frame}{Data Management Requirements}
\begin{overprint}
\onslide<1>
\begin{block}{Data capturing}
\begin{tabularx}{\linewidth}{p{35mm}X}
Flow Cytometry & FACS data \\
Histopathology & Zeiss slide images \\
Plant imaging & Lemnatec images, Flourogroscan images, 3D imaging \\
Infrared imaging & FLIR images \\
Chemical measurements & Chlorophyll content, Stomatal conductance \\
Visual observation & Manual reports (plant, mouse phenotypes)\\
$\cdots$ & $\cdots$
\end{tabularx}
\end{block}

\onslide<2>
\begin{block}{Metadata capturing}
\begin{tabularx}{\linewidth}{p{30mm}X}
Project & Project proposal, project plan \\
Investigation & Objectives, design \\
Materials & Lines/genotypes, samples, growth conditions \\
Devices & Specs, settings, versions \\
Processes & Workflows, protocols, variations \\
Measurements & Data, images \\
Analysis & Observations, results \\
$\cdots$ & $\cdots$
\end{tabularx}
\end{block}

\onslide<3>
\begin{block}{Data management tasks}
\begin{itemize}
\item Data distribution \& sharing
\item Data publishing
\item Access control
\item Archival \& versioning
%\item Structure \& metadata
\item Data discovery \& analysis
\item Data integration
\end{itemize}
\end{block}
\end{overprint}
\end{frame}

One of the requirements of PODD is to support data capturing from
various platforms.

As you can see on this slide. there are quite a number of different
ways of imaging and measuring a subject. The challenge here is,
besides capturing the raw image files and observations, capture
sufficient metadata.

Another requirement is managing the metadata. Metadata can be
generated on different levels and at different times. All this
information needs to be recorded in the repository together with
the data.

The above requirements span across the whole range of data management
tasks, including distribution, publishing, access control and versioning. 
In the phenomics domain, the development of a repository that supports 
these tasks need to take diversity and change into consideration.

\begin{frame}{PODD: an ontology-driven repository}
\begin{block}{Goals}
    \begin{itemize}
        \item Acquisition and storage of large volumes of data
        \begin{itemize}
        	%\item Different platforms \& different formats
        	\item Distribution, access control, versioning, etc.
	    \end{itemize}
        \item Data conxtualization
        \begin{itemize}
			\item Logical organization
        	\item Provenance tracking
			\item Discovery \& integration
        \end{itemize}
        \item Prepare for change
        \begin{itemize}
        	\item Changes in domain model
        \end{itemize}
    \end{itemize}
\end{block}
\pause
\begin{block}{Approach}
    \begin{itemize}
        \item An ontology-driven approach
        \item \textbf{Ontologies as the domain model}
        \item Benefits: flexibility \& extensibility
    \end{itemize}
\end{block}
\end{frame}

@@

Ontologies are a formal representation of a particular application
domain. Besides being a controlled vocabulary that define important
concepts and relations, an ontology can also define complex constraints
on these concepts and relations, making it a very expressive
conceptualization of the domain.

@@

@@

\section{Related Work}
\begin{frame}{Related Work}
\begin{block}{FuGe -- Functional Genomics Experiment}
    \begin{itemize}
        \item \emph{Material}, \emph{Protocol}, \emph{Data}, etc.
        \item Can be extended to support phenomics
        \item[$\cross$] Defined in UML \& mapped to database schemas -- difficult to extend for new concepts
    \end{itemize}
\end{block}
\pause
\begin{block}{OBI -- Ontology for Biomedical Investigations}
    \begin{itemize}
        \item \textit{``An integrated ontology for the description of life-science and clinical
        investigations.''}
        \item Comprehensive: 2,600+ classes, 10,000+ axioms
        \item[$\cross$] Complex, computationally ($\mathcal{SHOIN}(D)$)
    \end{itemize}
\end{block}
\end{frame}

FuGe is a UML and database based conceptual model for functional
genomics experiments. It models important domain concepts such as
material, protocol and data and how they relate to each other. 
The model is defined as UML class diagrams and mapped to database 
schemas, as a result, FuGe may be difficult to extend.

The OBI ontology - ontology for biomedical investigation is a
comprehensive but complex ontology. It defines domain concepts 
starting from a very high abstract level. As a result, it is quite 
big, containing about 26 hundred classes and 10 thousand logic
axioms. It is also computationally expensive. We need to performing 
reasoning repeatedly in PODD, we chose not to use OBI as our domain 
ontology.

\begin{frame}{Related Work}
\begin{block}{Web Ontology Language (OWL)}
\begin{itemize}
    \item Precise, open \& extensible -- exactly what we need!
    \item Provides core language constructs \& vocabularies for expressing complex ontologies -- \emph{data models}
    \item APIs, query engines \& automated reasoners available
\end{itemize}
\end{block}
\begin{block}{Fedora Commons}
\begin{itemize}
	\item Mature open-source digital repository software
	\item Modular \& extensible
	\item Widely used 
\end{itemize}
\end{block}
\end{frame}

@@

Fedora Commons is an open-source digital contents management system. It's designed
to be modular and extensible so that it can be used with a variety of plugins
and configurations. It's a very mature software and has been widely used 
to implement institution-wide repository systems.

\section{The PODD Ontology}
\begin{frame}{The PODD Ontology}
\begin{block}{Modeling essentials}
%    \begin{itemize}
%    \item Domain entities
%        \begin{itemize}
%        \item Abstract concepts
%        \item Concrete objects
%        \end{itemize}
%    \item Domain entities defined using OWL \emph{ontologies}
%    \end{itemize}
%\end{block}
%\pause
%\begin{block}{Modeling in OWL}
    \begin{itemize}
    \item Domain concepts -- \emph{OWL classes}
    \item Inter-concept relations -- \emph{OWL predicates} \& \emph{OWL restrictions}
    \item Concrete domain objects -- \emph{OWL Individuals}
    \item Comments, descriptions -- \emph{OWL annotations}
    \end{itemize}
\end{block}
\pause
\begin{block}{Benefits}
	\begin{itemize}
	\item Extensibility through inheritance
	\item Reuse \& integration through \textbf{ontology mapping} \& \textbf{ontology annotation}
		\begin{itemize}
		\item Gene Ontology, Plant Ontology, etc.
		\end{itemize}
	\end{itemize}
\end{block}
\end{frame}

%\begin{frame}{The Ontology-driven Approach}
%\begin{block}{Benefits}
%    \begin{itemize}
%    \setlength{\itemsep}{6pt}
%    \item<1-> Greater extensibility
%        \begin{itemize}
%        \item New concepts/relations can be easily added/modified
%        \end{itemize}
%    \item<2-> Better reuse \& integration
%        \begin{itemize}
%        \item Ontologies are \emph{open}
%        \item Other ontologies can be integrated
%        \begin{itemize}
%        	\item Class, property \& individual \textbf{mapping}
%			\item Ontological \textbf{annotations}
%        \end{itemize}
%        \end{itemize}
%    \item<3-> Balance between expressivity \& reasoning complexity
%        \begin{itemize}
%        \item Formal semantics enables automated query \& analysis
%        %\item Off-the-shelf tools available
%        \end{itemize}
%    \end{itemize}
%\end{block}
%\end{frame}

@@

Firstly, the repository can be more extensible. In PODD abstract
domain concepts are stored in the repository as well. After we've 
defined the domain ontology, concepts and properties can be extended 
to create new ones. In such a way new concepts and relations
can be added more easily than if the information is stored in a database.

Note here that concept and object definitions are all versioned so
we can freely modify an concept without affecting existing objects.

@@

\begin{frame}{The PODD Ontology -- Overview}
\vspace{-12pt}
 \begin{figure}[t]
  \begin{center}
  \includegraphics[trim=48mm 30mm 38mm 20mm, clip, width=112mm]{model.pdf}
   \end{center}
  \label{fig:ont}
 \end{figure}
\end{frame}

In this slide we show how some high-level concepts are logically
connected in the PODD domain model. Project is the top-level concept
with a bunch of attributes. It is also connected to some other
concepts like ProjectPlan, Platform and Investigation. Each of these
concepts also should have a number of core metadata attributes.

The concepts here are organized in a tree-like structure. What is not
shown on this slide is cross references between concepts. They are
useful when a particular object is referred to by some other objects.

\begin{frame}{The PODD Ontology -- An Example}
\begin{example}{The \textbf{\emph{Project}} concept}
    \begin{itemize}
    \item The top-level concept
    \item Constraints on inter-object relations \& attributes
    \end{itemize}
    \pause
    {\small
    \begin{align*}
    Project &\sqsubseteq~ =~ 1~ hasProjectPlan \sqcap \forall hasProjectPlan.ProjectPlan\\
            &\sqsubseteq~ \geq~ 1~ hasInvestigation \sqcap \forall hasInvestigation.Investigation\\
            &\sqsubseteq~ =~ 1~ hasStartDate \sqcap \forall hasStartDate.\textit{xsd:date}\\
            &\sqsubseteq~ \leq~ 1~ hasPublicationDate \sqcap \forall hasPublicationDate.date
    \end{align*}}
\end{example}
\begin{block}{}
    \begin{itemize}
    \item \textcolor{blue}{\textbf{Extensibility}} from inheritance of OWL classes \& predicates
    \end{itemize}
\end{block}
\end{frame}

In this slide I show an incomplete modeling of the top-level concept
in our ontology, the Project. In our model of phenomics, everything starts 
from a research project and all other objects, such as Investigation and
Material are child or descendant objects.

As you see, we define constraints on attributes and inter-concept relations
of the Project concept. This is how we model all of our concepts.

@@

%\begin{frame}{The PODD Ontology}
%\onslide<2->
%\begin{columns}[t]
%\column{.5\textwidth}
%\begin{itemize}
%    \item Inspired by FuGe \& OBI
%    \item Aim: define all \emph{essential} domain concepts, attributes \& relations
%    \item $\sim30$ classes, \\$\sim 80$ predicates, \\$\sim 200$ Axioms
%    \item Faster reasoning, querying, etc.
%\end{itemize}
%\vfill
%\onslide<1->
%\column{.5\textwidth}
%\vspace{-10mm}
% \begin{figure}[t]
%  \begin{center}
%  \includegraphics[trim=3mm 26mm 2mm 8mm, clip, height=72mm]{hier.pdf}
%   \end{center}
% \end{figure}
%\end{columns}
%\end{frame}

@@

\begin{frame}{PODD Ontology -- Roles}
\begin{block}{Ontologies \textbf{\emph{drive}} repository functions}
\begin{list}{}{}
\item[\textcolor{blue}{\textbf{Presentation}}]~\\% Drives the rendering of the web pages
    \begin{itemize}
    \item Object creation, editing, display, etc.
    \end{itemize}
\item[\textcolor{blue}{\textbf{Storage}}] ~\\
	\begin{itemize}
	\item Object (de)serialization to/from ontologies
	\end{itemize}
\item[\textcolor{blue}{\textbf{Validation}}] ~\\%Ontology reasoning performed
    \begin{itemize}
    \item Validation based on concept constraints
    \end{itemize}
\item[\textcolor{blue}{\textbf{Discovery}}] ~\\%Multiple ways of finding information
    \begin{itemize}
    \item Queries using SPARQL
    \item Full-text search
    \end{itemize}
\end{list}
\end{block}
\end{frame}

@@

\begin{frame}{The PODD Repository: The High-level Architecture}
\vspace{-10pt}
 \begin{figure}[t]
  \begin{center}
  \includegraphics[trim=48mm 30mm 38mm 20mm, clip, width=96mm]{architecture.pdf}
   \end{center}
  \label{fig:arch}
 \end{figure}
\end{frame}

Here you see the high-level architecture of the PODD repository. 
We take a layered approach in designing and developing the system.

The data access layer at the bottom is responsible for 
serializing and deserializing concepts and objects in 
the underlying storage. Objects and their metadata are
stored in Fedora Commons, a very widely-used digital
contents repository software system. iRODS is a federated 
data storage software that can be plugged into Fedora
Commons as a storage module. We use iRODS to store the
Fedora datastreams.

Object definitions in RDF will be stored in a Sesame triple 
store and a Lucene index so that we can query and search for
information. We also use databases to store information
about users, access rights and some other things.

The business logic layer is responsible for managing the 
lifecycles and interactions of objects and concepts. OWL
reasoning is a central component of this layer as we use
ontologies in all sorts of data management tasks.

On top of that is the security layer, which handles authentication
and authorization. We have a fine-grained authorization module
where access rights are granted on a per-user and per-project basis.

At the top of the stack is the interface layer. PODD is a web-based
system so all functionality can be accessed through a browser as
well as through API calls.

\section{Conclusion}
\subsection*{ }
\begin{frame}{Conclusion}
\begin{block}{To recap}
\begin{itemize}
\item Large amounts of data need to be managed
    \begin{itemize}
    \item There is a need for data archival, storage \& discovery
    \end{itemize}
\item Current approaches lacking/inadequate/inflexible
    \begin{itemize}
    \item Emerging processes, platforms, technologies require a extensible conceptual framework
    \end{itemize}
\item An ontology-driven architecture as the foundation of PODD
    \begin{itemize}
    \item Ontologies as the domain model
    \item Extensible \& open
    \end{itemize}
\end{itemize}
\end{block}
\end{frame}

As a result, in PODD we propose an ontology-driven architecture
for developing extensible data repositories. In this framework
databases take a back seat and ontologies are used to describe
the domain. We believe it is more extensible, flexible and open.

\begin{frame}{Conclusion}
\begin{block}{Where we are now}
\begin{itemize}
\item PODD ontology for phenomics research
\item Development of basic repository functionality
\item Development of PODD web interface
\end{itemize}
\end{block}
\pause
\begin{block}{What's next}
\begin{itemize}
\item Development of batch data import/export processes
\item Development of object discovery services
\item Integration with Shibboleth authentication 
\item \textcolor{gray}{Exposing data for discovery}
\item \textcolor{gray}{Integrating with other data sources}
\end{itemize}
\end{block}
\end{frame}

So far we have developed an ontology for plant and
mouse phenomics research. Centered around this ontology
we have implemented basic repository functionalities
including object storage, versioning, validation, 
publishing, searching and presentation generation.

In future we will implement a number of tasks that are
important for phenomics research. These include data
import and export, object discovery and more flexible
authentication with Shibboleth.

We will also be looking at how we can expose data in
the PODD repository to external world and how they 
can be integrated with existing datasets.

\begin{frame}
	\frametitle{}
    \begin{center}
\Huge \textsc{\textcolor{blue}{\textbf{Thank You!}}}
\end{center}
\vspace{+20pt}
\begin{block}{Acknowledgment}
Faith Davies, Gavin Kennedy, Jane Hunter @
eResearch Lab, School of ITEE, UQ
\end{block}
\end{frame}
